---
title: "Prepare data from the of the Exposome Challenge data set"
author: "Carles Criado Ninà"
output:
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    theme: cerulean
---

<style>
  <!-- To have nice indentation in toc_float. Copied from:
  https://github.com/rstudio/rmarkdown/issues/1909 -->
  .tocify-header {
    text-indent: initial;
  }
  .tocify-subheader > .tocify-item {
    text-indent: initial;
    padding-left: 2em;
  }
</style>

***

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r libaries, echo=FALSE}
library(Biobase)
```


# Load the data

First, we load the data sets and explore their structure.
The data is obtained from the 
[Exposome Data Challenge Event 2021](https://www.isglobal.org/-/exposome-data-analysis-challenge)
and is available at [github](https://github.com/isglobal-exposomeHub/ExposomeDataChallenge2021).
```{r load_data}
load("ExposomeChallengeStartingPoint.RData")
```

## Codebook (DataFrame)
```{r codebook}
str(codebook) |> str()
```

## Phenotype (DataFrame)
```{r phenotype}
phenotype |> str()
```

## Covariates (DataFrame)
```{r covariates}
covariates |> str()
```

## Exposome (DataFrame)
```{r exposome}
exposome |> str()
```

## Serum metabolome (ExpressionSet)
```{r serum}
exprs(metabol_serum) |> str()
pData(metabol_serum) |> str()
fData(metabol_serum) |> str()
```

## Urine metabolome (ExpressionSet)
```{r urine}
exprs(metabol_urine) |> str()
pData(metabol_urine) |> str()
fData(metabol_urine) |> str()
```

## Proteome (ExpressionSet)
```{r proteome}
exprs(proteome) |> str()
pData(proteome) |> str()
fData(proteome) |> str()
```


# Subset the observations present in all the datasets

The datasets have a different number of observations.
We want to keep only the observations present in all the datasets.
```{r build_index}
index <- (intersect(pData(metabol_serum)[,'ID'],
                    pData(metabol_urine)[,'ID']) |>
            intersect(pData(proteome)[,'ID']))
all(index %in% phenotype[,'ID'])
all(index %in% exposome[,'ID'])
all(index %in% covariates[,'ID'])
length(index)
```

Now we can subset all the datasets on the index.
```{r subset}
phenotype     <- phenotype[phenotype[ ,'ID'] %in% index, ]
exposome      <- exposome[exposome[ ,'ID'] %in% index, ]
covariates    <- covariates[covariates[ ,'ID'] %in% index, ]
metabol_serum <- metabol_serum[ ,index]
metabol_urine <- metabol_urine[ ,index]
proteome      <- proteome[ ,index]
```

# Save the data on CSV

Finally, we can save all the datasets on CSV files
so we can load them on Python.

```{r save}
directory <- "./datasets"
write.csv(codebook,   file.path(directory, "codebook.csv"))
write.csv(phenotype,  file.path(directory, "phenotype.csv"))
write.csv(exposome,   file.path(directory, "exposome.csv"))
write.csv(covariates, file.path(directory, "covariates.csv"))

write.csv(exprs(metabol_serum), file.path(directory, "metabol_serum_exprs.csv"))
write.csv(pData(metabol_serum), file.path(directory, "metabol_serum_pData.csv"))
write.csv(fData(metabol_serum), file.path(directory, "metabol_serum_fData.csv"))

write.csv(exprs(metabol_urine), file.path(directory, "metabol_urine_exprs.csv"))
write.csv(pData(metabol_urine), file.path(directory, "metabol_urine_pData.csv"))
write.csv(fData(metabol_urine), file.path(directory, "metabol_urine_fData.csv"))

write.csv(exprs(proteome), file.path(directory, "proteome_exprs.csv"))
write.csv(pData(proteome), file.path(directory, "proteome_pData.csv"))
write.csv(fData(proteome), file.path(directory, "proteome_fData.csv"))
```

***

```{r sessioninfo}
sessionInfo()
```