---
title: "Try the clValid package on the Exposome Challenge data set"
author: "Carles Criado Ninà"
output:
  html_document: 
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: true
    theme: cerulean
---

<style>
  <!-- To have nice indentation in toc_float. Copied from:
  https://github.com/rstudio/rmarkdown/issues/1909 -->
  .tocify-header {
    text-indent: initial;
  }
  .tocify-subheader > .tocify-item {
    text-indent: initial;
    padding-left: 2em;
  }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r libaries, echo=FALSE}
library(clValid)
```

***

# Introduction

[clValid](https://www.rdocumentation.org/packages/clValid/versions/0.7/topics/clValid)
is an R package
that provides a way to evaluate different clustering methods
on the same data.
It reports validation metrics
that can be chosen between three classes:
"internal", "stability", and "biological".

Using this package,
I will compare a set of clustering methods
over a range of cluster numbers,
which will become a grid of 8 x 10 "parameters" to explore.
Because of the size of the grid,
I choose to report only the "internal" metrics,
since they are the fastest to compute
and this study does not aim to be a profound exploration of the data.

First, I will evaluate the results over the metabolomics raw data.
Then, I will do a PCA analysis,
select the numbers of PCs that explain 80% of the variance,
and repeat the analysis on the learned features.
Finally, I will repeat the same two steps on the exposome data.


# Load the data

First, we load the data previously saved on CSV files.

```{r load_data}
directory <- "./ExposomeChallengeData/datasets"

# Exposome and metabolomics data (min-max normalized)
metabol_exprs <- read.csv(file.path(directory, "metabol_joint_exprs_minmax.csv"), row.names = 1)
exposome <- read.csv(file.path(directory, "exposome_minmax.csv"), row.names = 1)
```


# Metabolomics data

## Raw data

Try clValid on the raw data:

```{r clValid_metabol_raw}
CL_METHODS <- c("hierarchical", "kmeans", "diana", "som", "sota", "pam", "clara", "agnes")
N_CLUST <- 2:10
MAX_ITEMS <- nrow(metabol_exprs)

clval_metabol <- clValid(metabol_exprs,
                       nClust = N_CLUST,
                       clMethods = CL_METHODS,
                       maxitems = MAX_ITEMS,
                       validation = "internal",
                       metric = "euclidean",
                       method = "ward")

optimalScores(clval_metabol)
plot(clval_metabol)
```

No clustering method seem to find an optimal number of clusters.
Almost all of them seem to achieve better metrics
with a higher number of clusters.
I could use the elbow method to select an optimal number,
but it doesn't seem to be informative.


## PCA

Try clValid on the learned features (PCA):

```{r clValid_metabol_pca}
EXPLAINED_VARIANCE = 0.8

metabol_pca <- prcomp(metabol_exprs)
plot(metabol_pca)
num_pcs <- sum(summary(metabol_pca)$importance[3,] <= EXPLAINED_VARIANCE)
metabol_pca <- metabol_pca$x[,1:num_pcs]

clval_metabol_pca <- clValid(metabol_pca,
                       nClust = N_CLUST,
                       clMethods = CL_METHODS,
                       maxitems = MAX_ITEMS,
                       validation = "internal",
                       metric = "euclidean",
                       method = "ward")

optimalScores(clval_metabol_pca)
plot(clval_metabol_pca)
```

The PCA transformation doesn't seem to help,
the results are very similar.


## Try clValid on the exposome data

Try clValid on the raw data:

```{r clValid_exposome_raw}
clval_exposome <- clValid(exposome,
                       nClust = N_CLUST,
                       clMethods = CL_METHODS,
                       maxitems = MAX_ITEMS,
                       validation = "internal",
                       metric = "euclidean",
                       method = "ward")

optimalScores(clval_exposome)
plot(clval_exposome)
```

Here, I find that the majority of the methods
converge in an **optimal number of 6 clusters**,
which is confirms the results I found
with both *classic* clustering methods
and deep clustering methods (see Jupyter notebooks).


Try clValid on the learned features (PCA):

```{r clValid_exposome_pca}
exposome_pca <- prcomp(exposome)
plot(exposome_pca)
num_pcs <- sum(summary(exposome_pca)$importance[3,] <= EXPLAINED_VARIANCE)
exposome_pca <- exposome_pca$x[,1:num_pcs]

clval_exposome_pca <- clValid(exposome_pca,
                       nClust = N_CLUST,
                       clMethods = CL_METHODS,
                       maxitems = MAX_ITEMS,
                       validation = "internal",
                       metric = "euclidean",
                       method = "ward")

optimalScores(clval_exposome_pca)
plot(clval_exposome_pca)
```

The PCA converted data seem to reinforce the same result,
finding again an optimal number 6 clusters,
but with more pronounced differences.


# Discussion

The results obtained with this exploration
are in line with what I found on my earlier exploration
trying different clustering methods,
both *classic* methods and using deep learning
(see Jupyter notebooks).
The metabolomics data as provided
does not seem to lend itself to good clustering quality.

On the other hand the exposome data
shows a clear optimum number of 6 clusters
with almost all the methods explored.
On my earlier exploration
I found that this clusters overlap with the classes of the variable *cohorts*,
which suggests a that there is a strong batch effect on the data.



# SessionInfo

```{r sessioninfo}
sessionInfo()
```